A family of accessibility measures

### 2024-11-07 paper/manuscript.qmd
Draft of introduction and lit review, draft attempt at numerical formulation

ANAASTASIA: A brief intro describing the aims of the paper and anticipating the conclusions needs to go here.

<!-- the paper's aim-->
This work aims to re-unify accessibility literature with its Newtonian roots by mapping it onto the family of gravity models. Not only to clarify accessibility's interpretation as a measure, but also to encourage future developments that draw from its physics roots, gravity modelling siblings, and spatial interaction history. We began by tracing the development of spatial interaction from @ravensteinLawsMigration1889 to it's split into accessibility research from @hansen1959 to today. In the following section, we will derive different types of accessibility measures, following the family of gravity models namely: non-constrained, constrained, unconstrained/Total constrained, singly constrained, composite singly constrained, doubly constrained. Then, we will provide empirical examples of the types of measures. Finally, we'll discuss the use of each measure and broadly the implications of ignoring the issue of proportionality in accessibility analysis.

@@

<!--The inquiry into spatial interaction traces to Newts Universal Gravitation-->
The patterns of people's movement in space have been a subject of scientific inquiry for centuries. From as far back as 1858 with Carey's Principles of Social Science where he stated that "man [is] the molecule of society [and their interaction is subject to] the direct ratio of the mass and the inverse one of distance" [see McKean's -@mckeanManual1883 Manual of Social Science, pp. 37-38], investigations of how people interact in space has often been explicitly coloured by the features of Newton's Law of Universal Gravitation, as posited in his _Principia Mathematica_ of 1687 (see @eq-grav-prop):

@@

@eq-grav-prop is one of the most famous equations in all of science, and it states that the force of attraction $F$ between a pair of bodies $i$ and $j$ is directly _proportional_ to the product of their masses $M_i$ and $M_j$, and inversely _proportional_ to the square of the distance between them $D_{ij}$. Direct proportionality only means that as the product of the masses increases, so does the force. Likewise, inversely proportional only means that as the distance increases, the force decreases. To quantify the magnitude of the force, an empirical factor is required to convert the proportionality into an equation, which ensures that the force $F$ matches empirical observations of the force of attraction between masses. Ultimately, the equation for the force is as seen in @eq-grav:

@@

\noindent where $G$ is a proportionality constant obtained empirically and called _Cavendish gravitational constant_, after Henry Cavendish's experimental results of 1798--more than a century after Newton's _Principia_ was published--that brought it to within 1\% accuracy level.

<!--Using Newts Universal Gravitation: Ravenstein 1889 to Stewart 1948-->
Following Carey, research into human spatial interaction continued apace in different contexts. In 1885 and 1889 Ravenstein (who appears to have been unaware of Carey's treatise) proposed his "Laws of Migration", based on his empirical analysis of migration flows in various countries. In these works, Ravenstein posited a directly proportional relationship with the size of the destination (i.e., centres of commerce and industry), and an inversely proportional relationship between the size of flows and separation between origin and destination, in clear similitude to Newton's gravitational laws. Over time other researchers kept discovering, seemingly independently, some principles of human spatial interaction that resembled Newton's universal law of gravitation. For example, @reilly1929methods (who seems to have been unaware of Carey and Ravenstein) formulated a law of retail gravitation  <!--Rilley does not cite Newton, nor Carey, nor Ravenstein-->, expressed in terms of equal attraction to competing retail destinations. Later, Zipf (seemingly unaware of Carey, Ravenstein, and Rilley) proposed a $\frac{P_1P_2}{D}$ hypothesis for the case of goods movement by railways @zipfHypothesisCaseRailway1946, intercity personal movement @zipfHypothesisIntercityMovement1946,and information @zipfDeterminantsCirculationInformation1946. This hypothesis stated that the magnitude of flows was proportional to the product of the populations of settlements, and inversely proportional to the distance between them.

A common theme among these early researchers of human spatial interaction is that a proportionality constant similar to $G$ in @eq-grav is missing. Of the researchers cited here, only Reilly and Zipf expressed their hypotheses in mathematical terms. Reilly's hypothesis was expressed as follows:
$$
B_a = \frac{(P_a\cdot P_T)^N}{D_{aT}^n}
$${#eq-reilly}

\noindent where $B_a$ is the amount of business drawn to $a$ from $T$, $P_a$ and $P_T$ are the populations of the two settlements, and $D_{aT}$ is the distance between them. Quantity $N$ was chosen by Reilly in a somewhat _ad hoc_ fasion as 1, and he used empirical observations of shoppers to choose a value of $n = 2$.

@@

After Carey, it is in Stewart's [-@stewartDemographicGravitationEvidence1948] work on the principles of demographic gravitation that we find the strongest connection yet to Newton's law (Stewart was a physicist, whereas Ravenstein, Reilly, and Zipf were social scientists), besides awareness of preceding research, as he cites Reilly and Zipf as predecessors in the analysis of human spatial interaction. Stewart appears to have been the first author to express his theorized relationships for human spatial interaction with a proportionality constant $G$, as follows:
$$
F = G\frac{(N_1\mu_1)(N_2\mu_2)}{d_{12}^2}
$${#eq-stewart-force}

@@

- $F$ is the _demographic force_
- $N_1$ and $N_2$ are the numbers of people of in groups 1 and 2
- $\mu_1$ and $\mu_2$ are so-called _molecular weights_
- $d_{12}^2$ is the distance between $1$ and $2$
- And finally $G$, a constant that Stewart "left for future determination" [-@stewartDemographicGravitationEvidence1948, p. 34] but it is unclear whether he revisited

@@

Stewart's formulation of demographic force, developed in the context of what he called "social physics" [@stewartPrinciples1947], was problematic and had issues with inconsistent mathematical notation [e.g., in Stewart's 1948 paper, $G$ is used as a proportionality constant on p. 34 and again for the demographic energy on p. 54], but perhaps more seriously, it was permeated by a view of humans as particles and afflicted by unmitigated racism (Stewart assumed that the "molecular weight" $\mu$ was one for the average American, but "presumably...much less than one" for an Australian Aboriginal; p. 34). It did not take long for Stewart's ideas about "social physics" to fall out of favor among social scientists, but not before influencing Walter G. Hansen, then a student of city and regional planning at Massachusetts Institute of Technology.

<!--Hansen leans on Stewart's 1948 Population Potential measure. Two crucial things: 1) Stewart set the proportional constant to 1 *for future determination* and 2) Stewart suggests that keeping the exponent value of d_ij as 1 is sufficient. Hansen doesn't address the proportional constant of 1 explicitly in his application and picks an exponent other than 1 for d_ij as it fits empirically better: the issue of units are never addressed. -->
Hansen's day job was as an engineer with the the Bureau of Roads in the United States, and preoccupied with transportation's power to shape land uses. As such, he was neither scientist nor social scientist, but rather an applied researcher. Hansen's [-@hansen1959] paper in the Journal of the American Planning Association was instrumental in shedding some of the less convincing, or frankly objectionable, aspects of "social physics", to recast accessibility not as a model of physics, but as a model of human behavior useful to capture regularities in human mobility patterns. In his work, @hansen1959 borrowed the idea of a "population potential" of @stewartDemographicGravitationEvidence1948, which closely followed the Newtonian definition of energy potential, and applied it to the measurement of the potential of opportunities. The potential energy of one mass, say $M_i$, when applied to population instead of celestial bodies yields "potential population":
$$
S_{ij} = \sum_j O_j \cdot d_{ij}^\beta \\
$${#eq-accessibility}

In this equation, @hansen1959 replaced $M_j$ with _opportunities_ to derive _opportunity potential_, or more accurately, _potential of opportunities for interaction_. A more general form of this expression is as follows, to account for different impedance functions to reflect variations in travel behavior:
$$
S_{ij} = \sum_j O_j \cdot f(d_{ij})
$${#eq-accessibility-general}

$S_{ij}$ in @eq-accessibility is is a relative measure of the accessibility at $i$ to an activity located at $j$, $O_j$ is the size of the activity at $j$ and $d_{ij}$ is the distance or travel time between $i$ and $j$ with $\beta$ describing the effect size. Today, @hansen1959 is frequently cited as the father of modern accessibility analysis [e.g., @reggianiGuestEditorialNew2011] and Hansen-type accessibility is commonly referred to today as the gravity-based accessibility measure.

Of note, however is that between @stewartDemographicGravitationEvidence1948 and @hansen1959 the proportionality constant $G$ in @eq-grav vanished.

There is some evidence that @hansen1959 was aware of the import of this constant, since in his description of the process of interest he wrote that "the formulation states that the accessibility at point 1 to a particular type of activity at area 2 (say employment) is _directly proportional_ to the size of the activity at area 2 (number of jobs) and _inversely proportional_ to some function of the distance separating point 1 from area 2" (p. 73; our emphasis). The reason for this treatment of the proportionallity constant could simply be the empirical and theoretical limitations of the state of knowledge at the time when @hansen1959 drew from @stewartDemographicGravitationEvidence1948. After all, in the latter the proportionality constant was set to 1 and that _"the constant $G$ [was] left for future determination: a suitable choice of other units can reduce it to unity"_. In practice, this is still done today, and the proportionality constant continues to be implicitly set to $1$ even when the relationship is clearly of proportionality and not equality [see for instance the formula for accessibility at the top of Figure 1 in @wuUnifyingAccess2020]. Without a proportionality constant, however, the units of $S_ij$ are unclear: the unit of 'potential of opportunity for interaction' is left free to change as $\beta$ is calibrated for the population center with no discussion of a constant to balance the units.

<!--Accessibility from Hansen to today: very diverse, hard to interpret, and its lack of concern for the proportionality constant.  Our answer: the proportionality constant.-->
Since @hansen1959, accessibility analysis has been widely employed across planning disciplines, but to our knowledge, developments have not been concerned with explicitly determining a proportionality constant. The majority of studies develop on the breath of opportunity $O_j$ studied (i.e., work, shopping, schools, leisure () and their geographic case study selection, the accuracy of parameters that capture distance-decay $f(c_{ij})$, and additions to the opportunity $O_j$ attributes such as competitive demand considerations [@stoufferINTERVENINGOPPORTUNITIESCOMPETING1960; @weibull_axiomatic_1976; @shen1998; @allenMeasureCompetitiveAccess2020; @soukhovMultimodalSpatialAvailability2024]. However, the difficulty in interpreting gravity-based accessibility's meaning as anything other than a relative index of potential interaction (higher accessibility/lower accessibility), plagues this type of analysis to the date, and contributes to resistance to its application in practice [REFERENCE MISSING??]. The quantity and variations on @eq-accessibility-general compounds the issue of interpretability across studies. A testament to this inconsistent interpretation are the unsettled classifications generated by influential review works of the accessibility literature; for instance, @handyMeasuringAccessibilityExploration1997 classifies studies into gravity- and utility-based measures, @geursAccessibilityEvaluationLanduse2004 introduces further classification by expanding categories into potential- (or gravity), utility-, location-, and person- based measures with different components. @wuUnifyingAccess2020 demonstrate that most or all of these concepts can be unified into its main features: the relationship between travel-cost and corresponding reachable opportunities. Indeed, these authors demonstrate how previous classifications are special cases of travel-cost or opportunities constraints. From the initial Newtonian analogy, we interpret @wuUnifyingAccess2020's work as offering a positive contribution by circling back to re-unify mass $M$ and separation distance $D$; however, an important component is still missing—the proportionality constant $G$.

<!--
Anastasia: Maybe the idea in the text below is better left for the discussion to note the exceptions to the rule that the proportionality constant is implicitly set to one:

"To address our paper's aim: we refer to the gravity model of today (Wilson) as it is based in the same Newtonian gravitational analogy as accessibility as well as Paez et al. 2019 and Soukhov et al. 2023,2024"
-->

The aim of this paper's focus is on the proportionality constant. We wish to demonstrate the importance of transiting from proportionality to equality in @eq-accessibility-general, in order to balance the units of accessibility into more interpretable values with clear meanings. To address this matter we need not reinvent the wheel: instead we refer to the spatial interaction model as developed by @wilson1971. In essence, in what follows, we reunite the accessibility indicators literature with the spatial interaction literature as developed by @wilson1971 based on entropy maximization concepts. Although Wilson's approach is based on a different conceptual foundation than the old reference to Newtonian gravity, the work succeeded at identifying the steps from proportionality to equality to yield variations of proportionality constants: including the one that eluded @stewartDemographicGravitationEvidence1948 and has been ignored in almost all subsequent accessibility research.

# Enter Wilson and entropy maximization

The gravity model traces back to these early 20th century empirical investigations [@carrothersHistoricalReviewGravity1956; @griggRavensteinLawsMigration1977; @haynesGravitySpatialInteraction1985; @oshanSpatialStructureDebate2021]; it also retains the proportionality constant and the resulting output is a tangible unit.

$$
T_{ij} = k P_i^{\lambda} P_j^{\alpha} d_{ij}^{\beta}
$${#eq-4}

<!--Explain the features of the gravity model of today (Wilson) -->
Where $T_{ij}$ represents a $n$ x $m$ matrix of flows between $n$ origins $i$ and $m$ destinations $j$, $P$ are a vector of mass attributes at origin $i$ and $j$ and cost to overcome physical separation, and $d_{ij}$ is a $n$ x $m$ matrix. $d_{ij}$ is also frequently replaced by $f(c_{ij})$ in the literature, some function of $c_{ij}$ which decreases as $c_{ij}$ increases. Furthermore, the exponents are introduced to modulate the effect size of the variables, where $\lambda$ and $\alpha$ modulate $P$ and $\beta$ modulates $d$; when variables $T$, $P$ and $d$ are known, the exponents can be estimated to summarize the variable's effect size contribution to the known flows in the system. Finally and of most important note in this paper, $k$ is a scale parameter that makes the overall equation proportional to the rate characteristic of the modeled phenomenon. In other words, $T_{ij}$ explicitly measures _interaction_ as a unit of trips and the proportionality constant $k$ ensures the total sum of $T_{ij}$ represents the total flows in the data.

@@

>>>>

<!-- SUMMARY: Accessibility is related to mobility, however, institutionalized mobility-based transportation planning practice has treated accessibility as it's proxy. This has had problematic consequences. Conceptually, accessibility is a more complete and comprehensive evolution of mobility. As such, there has been recognition that the adoption of accessibility planning approaches are needed; however there are 1) too many definitions and 2) uncertainty in meaning of the outputs--> 
Accessibility is conceptually related to human mobility, it is often defined as the "potential of opportunities for interaction" [@hansen1959], opportunities such as places, events and social contacts that are important to people [@ferreiraReenactingMobilityAccessibility2020]. Accessibility adds a more comprehensive and complete understanding onto mobility, which simplify emphasizes the ability to move. However, the motivation of transportation planning has historically centered mobility and treated accessibility as its proxy, leading to problematic consequences [@handy2020]. In the rise of automobility-dominance, recent decades of mobility-based practice (e.g., highway expansion) have been marked by increased travel cost (time, effort, energy, financial cost) often not having an impact on accessibility levels [@handyACCESSIBILITYVSMOBILITYENHANCING2002]. In response, transportation planning practice has begin to adopt accessibility planning approaches, be it through the use of accessibility planning tools [@silvaAccessibilityInstrumentsPlanning2017] or through the traditional four-step transportation demand model [@ortuzar_2011_modelling]. However, institutional and conceptual barriers remain, many of which relate to the diversity of accessibility definitions [@handy2020] and tangible meaning of the outputs [@ferreiraReenactingMobilityAccessibility2020].

<!-- Considering population-level methods in transportation and city planning research, accessibility measures have been used to tackle a wide variety of research questions. They have taken on numerous definitions and analytical expressions; from simpler cumulative opportunities measures, to measures considering more complex travel behaviour and trip weighting characteristics, measures of _potential of opportunities for interaction_ are varied. Summarizing these definitions across studies have been challenging; a testament to these unsettled classifications are present in influential accessibility review literature [@handyMeasuringAccessibilityExploration1997; @geursAccessibilityEvaluationLanduse2004; @wuUnifyingAccess2020; @millerAccessibilityMeasurementApplication2018]. -->

<!-- PAPER'S AIM: improve interpretation of accessibility measure outputs. We do this by, 1) assuming accessibility is more complete understanding of mobiltiy, and connected them both mobility and accessibility are both based in the Netownian gravitational anology, lets introduce balancing factors like those  Gravity Model.   -->
In adopting the perspective that accessibility is an evolution of mobility conceptually, this paper aims to re-unify accessibility measures with its gravity-based Newtonian roots. It proposes to do so by mapping accessibility onto the family of gravity models. Not only to clarify accessibility's interpretation as a measure, but also to encourage future developments that draw from its physics roots, gravity modelling siblings, and spatial interaction history. We began by tracing the development of spatial interaction from @ravensteinLawsMigration1889 to it's split into accessibility research from @hansen1959 to today. In the following section, we will derive different types of accessibility measures, following the family of gravity models namely: non-constrained, constrained, unconstrained/Total constrained, singly constrained, composite singly constrained, doubly constrained. Then, we will provide empirical examples of the types of measures. Finally, we'll discuss the use of each measure and broadly the implications of ignoring the issue of proportionality in accessibility analysis.

@@

## The inquiry into spatial interaction traces to Newton's Universal Gravitation
The patterns of people's movement in space have been a subject of scientific inquiry for centuries. From as far back as 1858 in Henry C. Carey's _Principles of Social Science_ where "man [is] the molecule of society [and their interaction is subject to] the direct ratio of the mass and the inverse one of distance" [@mckeanManual1883, pp. 37-38], investigations of how people interact in space has often been explicitly coloured by the features of Newton's Law of Universal Gravitation [posited in _Principia Mathematica_ of 1687 and expressed in @eq-grav-prop].

@@

@eq-grav-prop is one of the most famous equations in all of science, and it states that the force of attraction $F$ between a pair of bodies $i$ and $j$ is directly _proportional_ to the product of their masses $M_i$ and $M_j$, and inversely _proportional_ to the square of the distance between them $D_{ij}$. Direct proportionality only means that as the product of the masses increases, so does the force. Likewise, inversely proportional only means that as the distance increases, the force decreases. To quantify the magnitude of the force, an empirical factor is required to convert the proportionality into an equation, which ensures that the force $F$ matches empirical observations of the force of attraction between masses. Ultimately, the equation for the force is as seen in @eq-grav, where $G$ is a proportionality constant obtained empirically:

@@

## Early human spatial interaction research: from Ravenstein 1889 to Stewart 1948
<!-- SUMMARY: from the 1880s to the 1940s, researchers have theoretically and empirically attempted to characterise human spatial interaction as some force of attraction $F$ that is directly proportion to the masses $Mi$ and $Mj$ and inversely proportional by their separation distance. This concept was captured with different expressions, but all tie back to the same Newtonian gravity analogy. -->
Following Carey, research into human spatial interaction continued in different contexts. In 1885 and 1889, Ravenstein proposed his "Laws of Migration", based on his empirical analysis of migration flows in various countries [@ravensteinLawsMigration1885;@ravensteinLawsMigration1889]. In these works, Ravenstein posited a directly proportional relationship with the size of the destination (i.e., centres of commerce and industry), and an inversely proportional relationship between the size of flows and separation between origin and destination. These propositions are similar to Newton's gravitational laws. Over time, other researchers discovered similar relationships. <!--Rilley does not cite Newton, nor Carey, nor Ravenstein, but I removed this curiosity as it's not central to our point here--> For example, @reilly1929methods formulated a law of retail gravitation, expressed in terms of equal attraction to competing retail destinations. Later, Zipf proposed a $\frac{P_1P_2}{D}$ hypothesis for the case of goods movement by railways @zipfHypothesisCaseRailway1946, intercity personal movement @zipfHypothesisIntercityMovement1946,and information @zipfDeterminantsCirculationInformation1946. The $\frac{P_1P_2}{D}$ hypothesis stated that the magnitude of flows was proportional to the product of the populations of settlements, and inversely proportional to the distance between them.

A common theme among these early researchers of human spatial interaction is that a proportionality constant similar to $G$ in @eq-grav is missing. Of the researchers cited, only Reilly and Zipf expressed their hypotheses in mathematical terms. Reilly's hypothesis was expressed as follows:
$$
B_a = \frac{(P_a\cdot P_T)^N}{D_{aT}^n}
$${#eq-reilly}

\noindent where $B_a$ is the amount of business drawn to $a$ from $T$, $P_a$ and $P_T$ are the populations of the two settlements, and $D_{aT}$ is the distance between them. Quantity $N$ was chosen by Reilly in a somewhat _ad hoc_ fashion as 1, and he used empirical observations of shoppers to choose a value of $n = 2$.

@@

After Carey, it is in Stewart's [-@stewartDemographicGravitationEvidence1948] work on the principles of demographic gravitation that we find the strongest connection yet to Newton's law. We suspect that may relate to academic background; Stewart was a physicist, whereas Ravenstein, Reilly, and Zipf were social scientists. Besides awareness of preceding research, as he cites Reilly and Zipf as predecessors in the analysis of human spatial interaction, Stewart appears to have been the first author to express his theorized relationships for human spatial interaction with a proportionality constant $G$, as follows:

$$
F = G\frac{(N_1\mu_1)(N_2\mu_2)}{d_{12}^2}
$${#eq-stewart-force}

@@

- $F$ is the _demographic force_
- $N_1$ and $N_2$ are the numbers of people of in groups 1 and 2
- $\mu_1$ and $\mu_2$ are so-called _molecular weights_
- $d_{12}^2$ is the distance between $1$ and $2$
- And finally $G$, a constant that Stewart "left for future determination" [-@stewartDemographicGravitationEvidence1948, p. 34]

@@

Stewart's formulation of demographic force, developed in the context of what he called "social physics" [@stewartPrinciples1947], was problematic. It had issues with inconsistent mathematical notation. For example, e.g., in @stewartDemographicGravitationEvidence1948, $G$ is used as a proportionality constant for both _demographic force_ and  _demographic energy_ on p. 34 <!--Antonio, is this what you meant?-->. And perhaps more seriously, it was permeated by a view of humans as particles, following physical laws, laws of which were afflicted by unmitigated racism. For instance, molecular weight $\mu$ was taken as one for the average American, but "presumably...much less than one....for an Australian aborigine" [p. 35]. Stewart's ideas about "social physics" fell out of favour among social scientists, but not before influencing the field of accessibility today.

In summary: from the 1880s to the 1940s, researchers have theoretically and empirically attempted to characterise human spatial interaction as some force of attraction $F$ that is directly proportion to the masses $M$ of a pair of locations ($i$ and $j$) and inversely proportional by some sort of separation distance $D_{ij}$. Those that captured the relationship empirically, approached the concept with different expressions, some including a proportionality constant and some not, but all tying back to the Newtonian gravity analogy.

## Hansen's gravity-based accessibility
<!-- SUMMARY: in Hansen's use of @stewartDemographicGravitationEvidence1948's Population Potential measure, we highlight two crucial omissions. 1) Stewart set the proportional constant $G$ to 1 _for future determination_, this constant is not explicitly addressed in @hansen1959 and accessibility continues to evolve without it. 2) Stewart suggests that keeping the exponent value of $d_ij$ as 1 is sufficient. In @hansen1959, an exponent other than 1 for $d_ij$ is selected based on empirical findings, however, the issue of _opportunity potential_ units are never addressed. -->

From @stewardDemographicGravitationEvidence1948, we arrive to Walter G. Hansen, the author of @hansen1959 which has been exceptionally influential in the accessibility literature. In this seminal paper @hansen1959, "accessibility is defined as the potential of opportunities for interaction" [p.73], it is "a generalization of the population-over-distance relationship or "population potential" concept developed by Stewart [@stewardDemographicGravitationEvidence1948]" [p.73], explicitly defined "as the potential of opportunities for interaction" [p.73]. As well as being a student of city and regional planning at Massachusetts Institute of Technology, Hansen was also an engineer with the the Bureau of Roads likely preoccupied with applied research that can be used to qualify transportation's power to shape land uses. @hansen1959 focused on @stewartDemographicGravitationEvidence1948's _population potential_ (expressed in (@eq-stewart-population-potential-sum)), leaving out Stewart's other formulaic contributions and objectionable aspects of "social physics". @hansen1959 recast @eq-stewart-population-potential-sum to reflect accessibility, a model of human behaviour useful in capturing regularities in mobility patterns . In this equation, @hansen1959 replaced $M_j$ with _opportunities_ to derive _opportunity potential_, or more accurately, _potential of opportunities for interaction_ as follows:
$$
S_{i} = \sum_j O_j \cdot d_{ij}^\beta \\
$${#eq-accessibility}

Alternatively, a more general form of this expression to account for different impedance functions:
$$
S_{i} = \sum_j O_j \cdot f(d_{ij})
$${#eq-accessibility-general}

$S_{i}$ in @eq-accessibility is is a relative measure of the accessibility at $i$, $O_j$ is the size of the activity at $j$ and $d_{ij}$ is the distance or travel time between $i$ and $j$ with $\beta$ describing the effect size. Today, Hansen is is frequently cited as the father of modern accessibility analysis due to his work in @hansen1959 [e.g., @reggianiGuestEditorialNew2011], and the Hansen-type accessibility is commonly referred to today as the gravity-based accessibility measure.

Of note, however is that between @stewartDemographicGravitationEvidence1948 and @hansen1959 the proportionality constant $G$ in @eq-grav vanished.  There is some evidence that @hansen1959 was aware of the importance of this constant as he wrote about _direct_ and _inversely proportional_ relationships between opportunities in one area, opportunities at another, and their separation distance. Furthermore, those reading @hansen1959 could consult @stewartDemographicGravitationEvidence1948 in which the proportionality constant $G$  was set to 1 and that "$G$ [was] left for future determination: a suitable choice of other units can reduce it to unity".

## From Hansen-type accessibility to accessibility today
<!-- SUMMARY: accessibility from Hansen to today is very diverse and has a lack of concern for the interpretability of output units; likewise, a lack of concern for balancing units such as a proportionality constant.  Our answer: let's focus on those omitted (or implicit set to 1) proportionality constants.-->

Since @hansen1959, accessibility analysis has been widely employed across planning disciplines, but to our knowledge, developments have not been concerned with explicitly determining a proportionality constant $G$. $G$ continues to be implicitly set to $1$ even when the fundamental relationship in accessibility is proportionality (e.g., $ S_{i}  \propto G \sum_j O_j \cdot f(d_{ij})$ and not equality [for instance, see the formula for accessibility at the top of Figure 1 in @wuUnifyingAccess2020]. Without a proportionality constant, however, the units of $S_i$ remain unclear: the unit of 'potential of opportunity for interaction' is left free to change as $\beta$ is calibrated for the population center with no discussion of a constant to balance the units.

To overview recent accessibility literature development, the majority of studies develop on; the breath and geographic case study selection of the opportunity $O_j$ studied and potentially impacted population []; the accuracy of parameters that capture distance-decay $f(c_{ij})$ []; and additions to the opportunity $O_j$ attributes such as competitive demand considerations [@stoufferINTERVENINGOPPORTUNITIESCOMPETING1960; @weibull_axiomatic_1976; @shen1998; @allenMeasureCompetitiveAccess2020; @soukhovMultimodalSpatialAvailability2024]. Overall, due to the diverse range of questions accessibility measures have been applied to help clarify, a plethora of variations in @eq-accessibility-general are present in the literature.

However, it has been raised that there is a difficulty in interpreting gravity-based accessibility's meaning as anything other than a relative index of potential interaction (higher accessibility/lower accessibility) []. This interpretation plagues this type of analysis, as implications beyond high-or-low access hotspots are often desired. What does the accessibility value mean? Difficultly in interpreting the output's tangible meaning contributes to resistance to its application in practice [@ferreiraReenactingMobilityAccessibility2020]. As reviewed, quantity and variations on @eq-accessibility-general also compounds the issue of interpretability across studies.

A testament to this inconsistent interpretation are the unsettled classifications generated by influential review works of the accessibility literature; for instance, @handyMeasuringAccessibilityExploration1997 classifies studies into gravity- and utility-based measures, @geursAccessibilityEvaluationLanduse2004 introduces further classification by expanding categories into potential- (or gravity), utility-, location-, and person- based measures with different components. @wuUnifyingAccess2020 demonstrate that most or all of these concepts can be unified into its main features: the relationship between travel-cost and corresponding reachable opportunities. Indeed, these authors demonstrate how previous classifications are special cases of travel-cost or opportunities constraints. From the initial Newtonian analogy, we interpret @wuUnifyingAccess2020's work as offering a positive contribution by circling back to re-unify mass $M$ and separation distance $D$; however, an important component is still missing, the proportionality constant $G$.

The aim of this paper's focus is on the proportionality constant. We wish to demonstrate the importance of transiting from proportionality to equality in @eq-accessibility-general, in order to balance the units of accessibility into more interpretable values with clear meanings. To address this matter, we need not reinvent the wheel: instead we refer to the spatial interaction model as developed by @wilson1971. In essence, in what follows, we reunite the accessibility indicators literature with the spatial interaction literature as developed by @wilson1971 based on entropy maximization concepts. Although Wilson's approach is based on a different conceptual foundation than the old reference to Newtonian gravity, the work succeeded at identifying the steps from proportionality to equality to yield variations of proportionality constants: including the one that eluded @stewartDemographicGravitationEvidence1948 and has been ignored in almost all subsequent accessibility research.

# Gravity-based accessibility's sibling: the gravity model

Today's gravity-based accessibility measure (@eq-accessibility-general) share the same early 20th century Newtonian-based  origins as the gravity model, the most widely used model in the study of spatial interaction [@ortuzar_2011_modelling;@wilson1971]. However, in some ways the gravity model and spatial interaction continued to evolve independently from @eq-accessibility-general, the former being developed into formalized transportation demand modelling practice to 'predict-and-provide', and the later taken up to describe land-use and transportation interactions by geography and social science- adjacent researchers and planners. <!-- Antonio, this is my sense at this point..?-->

There are plenty of insights to be shared across these sibling branches; they are both spatial interaction models From the perspective of gravity-based accessibility measure, this paper's focus is on the proportionality constant of the classic gravity model, which takes the following form:

$$
T_{ij} = k O_i D_j f(c_{ij})
$${#eq-gravity-model}

<!--Explain the features of the gravity model of today (Wilson) -->
Where $T_{ij}$ represents a $n$ x $m$ matrix of flows between an $n$ number of origins $i$ and an $m$ number of destinations $j$, $O_i$ and $D_j$ is a vector of mass attributes representing origin $i$ and destination $j$ respectively, and $f(c_{ij})$ is a $n$ x $m$ matrix representing some function of travel cost $c_{ij}$ which reflects travel impedance. In other words, $T_{ij}$ explicitly measures _interaction_ as a unit of trips and the single proportionality constant $k$ ensures the total sum of $T_{ij}$ represents the total flows in the data. It should be noted that the early analogy with the gravitational law proved too simplistic, and improvements to the gravity model (@eq-gravity-model) are the use of total trip ends (number of originating trips from zone $i$ are $O_i$ and ending trips at zone $j$ are $D_j$) instead of total populations and separation distance calibrated by a power $n$ (such as $P_i$, $P_j$, and $n$ in @eq-zipf, @eq-reilly and @eq-stewart-force). Finally, and of most important note in this paper, $k$ is a scale parameter that makes the overall equation proportional to the rate characteristic of the modeled phenomenon.

Traditionally, the development of the gravity model has put an emphasis on the interpretability of the resulting units [@kirbyNormalizingFactorsGravity1970; @wilsonSTATISTICALTHEORYSPATIAL1967;@wilson1971]. As such, @eq-gravity-model can take account of any additional information that can be built into the model to constrain the _interaction_ variable $T_{ij}$. As mentioned, $T_ij$ is in the unit of trips, so additional knowledge can be the total trip flows $O_i$ originating at each $i$ and the total trip flows $D_j$ terminating at $j$. These pieces of information can constrain $T_{ij}$ as follows, ensuring it reflects this knowledge:

$$
\sum_j T_{ij} = O_i
$${#eq-constraint1-gravitymodel}

$$
\sum_i T_{ij} = D_j
$${#eq-constraint2-gravitymodel}

Using these constraints, the gravity model @eq-gravity-model is typically distinguished in four cases, cases that are mathematically and conceptually consistent, so they have continued to be useful in classifying the gravity model in trip distribution modelling to this day [@ortuzar_2011_modelling]:

- neither @eq-constraint1-gravitymodel or @eq-constraint2-gravitymodel hold -- the unconstrained case;
- @eq-constraint1-gravitymodel holds -- the production constrained case (origin-constrained, a singly-constrained case); 
- @eq-constraint2-gravitymodel holds -- the attraction constrained case (destination-constrained, a singly-constrained case); 
- @eq-constraint1-gravitymodel and @eq-constraint2-gravitymodel both hold simultaneously -- the production-attraction constrained case (doubly-constrained).

In each of these four cases, the single proportionality constant $k$ is replaced by two sets of balancing factors $A_i$ and $B_j$ to ensure the case-specific constraints are satisfied. For the doubly-constrained case:  
$$
T_{ij} = A_i O_i B_j D_j f(c_{ij})
$${#eq-doublyconstrained-gravitymodel}

And the balancing factors equal:
$$
A_i = \frac{1}{\sum_j B_j D_j f(c_{ij})}
$${#eq-bf-A-gravitymodel}

$$
B_j = \frac{1}{\sum_j A_i O_j f(c_{ij})}
$${#eq-bf-B-gravitymodel}

For the singly-constrained cases, origin-constrained case replaces $B_j$ with 1 for all $j$ and $A_i$ is @eq-bf-A-gravitymodel, and the the destination-constrained case replaces $A_i$ with 1 for all $i$ and $B_j$ is @eq-bf-B-gravitymodel. In a way, the singly-constrained and doubly-constrained cases are variants of the unconstrained case.

## Entropy maximising: the gravity-model as statistical averages
Instead of the Newtonian gravity analogy, entropy maximisation is a popular model building approach taken in the spatial interaction studies to generate the gravity model. Namely, it is mathetmatically regorouse and offers a sensible interpretation of the resulting mode as a statistical averaging procedure. This allows the measure to be effective in describing behaviour in large populations but less effective in smaller decision making units in the region of interest.

The basis of the approach is to accept that there are macro, meso, and micro states -- where all micro states that are consistent with our information about macro states are equally likely to occur. For the _interaction_ variable of trips $T_{ij}$, as shown in @wilson1971, the number of micro states $W\{{T_{ij}}\}$ within a meso state $T_{ij}$ can be expressed as:
$$ 
W\{{T_{ij}}\} = \frac{T!}{\prod_{ij}T_{ij}!}
$${#eq-micro-meso-states}
<!--I LEFT OUT THE DERIVATION... NOT SURE IF ITS NECESSARY? -->
Using the maximizing entropy approach, we must find a set of $T_{ij}$ that maximizes @eq-micro-meso-states subject to our sensible macro-state constraints @eq-constraint1-gravitymodel, @eq-constraint1-gravitymodel and the following on total cost $C$ constraint:
$$
\sum_{ij}T_{ij}c_{ij} = C
$${#eq-constraint3-gravitymodel}

The natural logarithm of @eq-micro-meso-states is maximized under @eq-constraint1-gravitymodel, @eq-constraint1-gravitymodel and @eq-constraint3-gravitymodel for mathematical convenience, as both $ln W$ and $T_{ij}$ share the same maximum [@wilson1971; @ortuzar_2011_modelling]. The result is:

$$
T_{ij} = A_i O_i B_j D_j exp(-\beta c_{ij})
$${#eq-maxentropy-gravitymodel}

Where,
$$
A_i = \frac{1}{\sum_j B_j D_j  \exp(- \beta c_{ij})}
$${#eq-bf-maxentropy-A-gravitymodel}

and
$$
B_j = \frac{1}{\sum_j A_i O_j  \exp(- \beta c_{ij})}
$${#eq-bf-maxentropy-B-gravitymodel}

Results of maximisation are identical to @eq-doublyconstrained-gravitymodel, @eq-bf-A-gravitymodel, and @eq-bf-B-gravitymodel except the general $f(c_{ij})$ is $exp(-\beta c_{ij})$. From this approach, the gravity model can be interpretted to say that given the total number of individual trips (micro level) statistically represented at origin and destination zones (meso level), the cost of travelling (micro level, represented at the meso level), and the fixed total cost of travelling in the region (number of trips and cost of travel at the macro level). There is then a most probable distribution of trips between zones (meso-states) that stasify all the macro-state constraints.

@@

# References
